---
title: "Tidy Data Class"
author: "Kaleb J. Frierson"
date: "2024-09-24"
output: 
  github_document: 
    toc: TRUE
---

# Class Notes

Tidy data: 

What is it? 

An example of non tidy data is when treatment a and treatment b are variables and there are counts of each instead of variable treatment with options a or b. 

Relational data: 
  
  Data spread across tables with defined relations. Variables used to define
  these relations are keys. 
  
  Inner joins looks for overlap. Outer joins has x and brings in everything from 
  y. They can be left, right, or full. 
  
  They're called relational because they relate to each other through at least     one variable. 
  
Key functions: 

pivot_longer, seperate (single column with many variables), pivot_wider from long table format, bind_rows, *_join: 

# Coding

## Libraries/Setup

```{r calling libraries}

library(tidyverse)
library(readxl)
library(haven)

```

```{r}
pulse_df= 
  read_sas("data_import_examples/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()

```

## Pivot Longer

This needs to go from wide to long format: 

```{r}
pulse_tidy_df = 
  pulse_df |> 
  pivot_longer (
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi_score"
  )

```

if you tell sas what the variable prefix is using names_prefix, it will remove the prefix from it so that answers are only the abbreviation/name specific to that value of the variable. 

```{r}
pulse_tidy_df = 
  pulse_df |> 
  pivot_longer (
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) |> 
  mutate(
    visit = replace(visit, visit == "bl", "00m")
  ) |> 
  relocate(id, visit)

```


Now that you know how to do that, you can really do all of this in one code chunk without making a duplicated dataset: 

```{r}
pulse_df= 
  read_sas("data_import_examples/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer (
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi_score",
    names_prefix = "bdi_score_"
  ) |> 
  mutate(
    visit = replace(visit, visit == "bl", "00m")
  ) |> 
  relocate(id, visit)

```

One more example: 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = gd0_weight:gd18_weight, 
    names_to = "gd_time", 
    values_to = "weight"
  ) |> 
  mutate(
    gd_time = case_match(
      gd_time, 
      "gd0_weight" ~ 0, 
      "gd18_weight" ~ 18
  ))
  
```
case match requires you to list all options and name what they will be. You can also add a bonus for missing. Case when is more flexible. 

## Pivot Wider(S)

Now lets make an analysis result table: 

```{r}
analysis_df = 
  tibble(
    group = c("treatment", "treatment", "control", "control"), 
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 10, 4.2, 5)
  )


```

Pivot wider for human readability and can format as table using knitr!!

```{r}
analysis_df |> 
  pivot_wider(
    names_from = time, 
    values_from = mean
  ) |> 
  knitr::kable()
```


Lets move into data that exists across multiple elements and work on bringing them together: 

## Bind Tables

```{r}
fellowship_ring= 
  read_excel("data_import_examples/LotR_Words.xlsx", range = "B3:D6") |> 
  mutate(movie = "fellowship_ring")

two_towers= 
  read_excel("data_import_examples/LotR_Words.xlsx", range = "F3:H6") |> 
  mutate(movie = "two_towers")

return_king= 
  read_excel("data_import_examples/LotR_Words.xlsx", range = "J3:L6") |> 
  mutate(movie = "return_king")

lotr_df = 
  bind_rows(fellowship_ring, two_towers, return_king) |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = female:male, 
    names_to = "sex", 
    values_to = "words"
  ) |> 
  relocate(movie) |> 
  mutate(race = str_to_lower(race))

```

## Join FAS Datasets

Import litters dataset: 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
  ) |> 
  separate(
    group, into = c("dose", "day_of_treatment"), sep = 3
  )
```

Import pups: 

```{r}
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex, 
      1 ~ "male", 
      2 ~ "female"
    )
  )
```

Join the datasets!

```{r}
fas_df = 
  left_join(pups_df, litters_df, by = "litter_number") |> 
  relocate(litter_number, dose, day_of_treatment)
```



